---
title: "STAT 417 Project 3"
output: 
  html_document:
    theme: flatly
author: "Nazimuddin Shaikh"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

### Markov Chains

A computer is shared by 2 users who send tasks to a computer remotely and work independently. At any minute, any connected user may disconnect with probability 0.5, and any disconnected user may connect with a new task with probability 0.2. Let $X(t)$ be the number of concurrent users at time $t$ (in minutes). This is a Markov chain with 3 states: 0, 1, and 2. The probability transition matrix can be calculated and it is

$$\begin{bmatrix} 0.64 & 0.32 & 0.04 \\ 0.40 & 0.50 & 0.10 \\ 0.25 & 0.50 & 0.25 \\ \end{bmatrix}$$
        
Generate 10000 transitions of the Markov chain. What percent of times do you find your generated Markov chain in each of its 3 states?

---

### Solution

Creating the matrix:

```{r}
P = matrix(c(.64, .4, .25, .32, .5, .5, .4, .1, .25), nrow=3)
```

We also load the library `expm` for matrix exponentiation using the `%^%` operator provided by the package.

```{r, include=FALSE}
library(expm)
```

Using the matrix to generate 10000 transitions. For the sake of simplicity, only the first 5 generations.

```{r}
P%^%1
P%^%2
P%^%3
P%^%4
P%^%5
```

If we were to generate a 10000 transitions, our matrix would approach infinity, like so:

```{r}
P%^%10000
```

What we are trying to find to answer the problem of this question, is really the _Steady State Distribution_. We can find it since we know that the sum of the distribution quantities for each state is equal to 1, that is: $$\pi_1 + \pi_2 + \pi_3 = 1$$

To do this in R, we use our matrix and make a tibble, then solve using a system of equations. 

```{r}
A = rbind(t(P - diag(3)), c(1, 1, 1))
b = c(0, 0, 0, 1)
cbind(A, b)
qr.solve(A, b) # our steady state distribution
```

We can now use this to answer the question. We find that our markov chain is in state 1 40.00% of the time, in state 2 38% of the time, and in state 3 22% of the time. Otherwise summarized in the table below:

```{r, include=FALSE}
state = c(1, 2, 3)
perc = c(40, 38, 22)
tableF = rbind(state, perc)
```

```{r}
tableF
```